% !TEX encoding = UTF-8
\documentclass[12pt,a4paper]{article}

% ===== Encoding & Language =====
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% ===== Math =====
\usepackage{amsmath, amssymb, amsfonts}

% ===== Graphics =====
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}

% ===== Tables =====
\usepackage{booktabs}

% ===== Layout =====
\usepackage{geometry}
\geometry{margin=1in}

% ===== Links =====
\usepackage{hyperref}

% ===== Bibliography =====
\usepackage[numbers,sort&compress]{natbib}

% ===== Optional (recommended for better formatting) =====
\usepackage{setspace}
\usepackage{titlesec}


\title{\textbf{Evaluating Retrieval-Augmented Generation for Medical Question Answering and Literature Summarization}}

\author{
\\
\textit{Estin Bejaia} 
}

\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Large Language Models (LLMs) are increasingly used for medical information support, but hallucinated or weakly grounded outputs remain a critical risk in healthcare contexts. This paper presents and evaluates a clinical GraphRAG framework for mental health question answering with explicit evidence grounding and safety-aware routing. The proposed pipeline builds a domain knowledge graph from authoritative sources (WHO, NIMH, and DSM-derived material) through reproducible stages: source ingestion, markdown-based segmentation, contextual enrichment, schema-constrained entity-relation extraction, graph cleaning, targeted repair, and community summarization.

At inference time, the system applies a safety-first decision layer, then performs graph-first retrieval for in-scope questions using condition lookup, reverse symptom lookup, comparison reasoning, and entity-centric relation exploration. Optional LLM synthesis is restricted to retrieved evidence. Responses include both citations and source chunks to support traceability and manual verification.

The resulting framework provides an end-to-end, auditable workflow for grounded clinical QA and establishes a practical baseline for future quantitative evaluation and hybrid retrieval extensions.
\end{abstract}

\section{Introduction}

Recent LLM systems can generate fluent medical answers, but fluency is not equivalent to correctness. In healthcare settings, unsupported statements can mislead users about symptoms, treatments, urgency, and risk, especially when users treat conversational responses as practical guidance. This makes factual grounding and safety behavior first-class system requirements.

Retrieval-Augmented Generation (RAG) addresses part of this challenge by conditioning generation on external evidence. However, in clinical domains, many questions are relational rather than purely semantic: users ask about condition-symptom links, differential diagnosis, prevalence statements, treatment suitability, and safety signals. A graph representation is well suited to these relation-centric queries because it explicitly stores typed entities and typed edges.

This paper documents and evaluates an implemented Clinical GraphRAG system centered on mental health QA. The system is not a toy prototype: it includes reproducible data ingestion, checkpointed graph construction, schema-constrained extraction, graph cleaning and repair stages, safety gating, and a production API. The runtime path is graph-first and checkpoint-backed, with explicit citations and source chunks returned in each answer. 

The practical goals of this work are:
\begin{itemize}
    \item reduce unsupported generations by constraining responses to retrieved graph evidence,
    \item preserve provenance through chunk-level citations and metadata,
    \item improve robustness with repair passes and regression testing,
    \item enforce a safety short-circuit for crisis-like queries.
\end{itemize}

This manuscript therefore focuses on an engineering-valid, auditable RAG pipeline and its observed system behavior, while keeping claims aligned with what is currently implemented.


\section{Dataset Construction and Processing}

\subsection{Data Sources}

We focus on the mental health domain and construct a domain-specific corpus from authoritative public sources instead of relying only on generic QA benchmarks. The current source set includes:
\begin{itemize}
    \item WHO and NIMH mental health pages and documents,
    \item DSM-derived reference material after dedicated markdown filtering.
\end{itemize}

Source definitions are maintained in a registry and converted into a provenance manifest that records authority, condition tag, URL, parser mode, and output markdown path.

\subsection{Data Extraction and Contextual Enrichment}

The ingestion layer uses format-specific extraction:
\begin{itemize}
    \item HTML sources are converted with \texttt{trafilatura},
    \item PDF sources are converted with \texttt{pymupdf4llm}, with fallback behavior when layout mode fails.
\end{itemize}

After ingestion, markdown documents are parsed into nodes using structural headers (via Markdown parsing), not arbitrary fixed-length splitting. Each node keeps source metadata (authority, condition, source label, section/header path, and file path), enabling later citation and auditing.

For contextual enrichment, an LLM generates a short context sentence per node (stored as \texttt{context\_tag} and prepended to node text). This enrichment stage is checkpointed and resumable, so interrupted runs can continue deterministically under compatible node hash, model, and prompt version.

\textbf{Example data sample:}
\begin{itemize}
    \item \textbf{Context tag:} ``This passage focuses on diagnostic features and symptom structure for a mental health condition.''
    \item \textbf{Text chunk:} an original section-paragraph segment with preserved source metadata.
\end{itemize}

\subsection{Preprocessing}

Preprocessing is designed for quality control before graph extraction:
\begin{itemize}
    \item \textbf{Document-level filtering:} DSM markdown is filtered to remove non-clinical administrative sections and low-value chunks.
    \item \textbf{Node integrity checks:} fingerprinting and checkpoint validation to avoid stale cache reuse.
    \item \textbf{Citation normalization:} standardized source labels and section-aware citation strings.
    \item \textbf{Resumable enrichment:} context-tag generation with periodic checkpoint snapshots.
\end{itemize}

This pipeline preserves clinical context while improving downstream extraction consistency.

\subsection{Knowledge Graph Construction}

From enriched nodes, a clinical knowledge graph is extracted and refined through multiple control layers:
\begin{itemize}
    \item \textbf{Schema-constrained extraction:} the LLM outputs typed entities and typed relations under a fixed allowlist.
    \item \textbf{Canonicalization and merge:} aliases and abbreviations are normalized (e.g., CBT $\rightarrow$ COGNITIVE BEHAVIORAL THERAPY), and equivalent entities are merged.
    \item \textbf{Validation and cleaning:} invalid triplets, placeholder entities, low-signal entities, and inconsistent edges are removed.
    \item \textbf{Repair stages:} targeted repair prompts and curated payload injection fill known factual gaps (e.g., prevalence and differential diagnosis coverage).
    \item \textbf{Community construction:} Leiden clustering produces graph communities; summaries are generated by LLM when configured, otherwise by heuristic fallback.
\end{itemize}

The extracted graph stores both relational structure and provenance maps (chunk citations and chunk payload metadata), enabling traceable answer construction at runtime.

All major artifacts are checkpointed (nodes, enriched nodes, graph state, communities, summaries), supporting reproducible rebuilds and resumable long runs.

\section{Problem Formulation}

\subsection{Task Definition}

Let $q$ denote a user query in the mental health domain, and let $h$ denote optional session history for follow-up context. The system must return a response that is both useful and auditable.

We define the returned payload as:
\begin{equation}
R = \left(A, \mathcal{C}, \mathcal{S}, t, b\right)
\end{equation}
where $A$ is the answer text, $\mathcal{C}$ is the citation list, $\mathcal{S}$ is the set of supporting source chunks, $t$ is the query type, and $b$ is a safety/status block (e.g., crisis or out-of-scope flags).

Formally:
\begin{equation}
A = f(q, h, \mathcal{G}, \Sigma)
\end{equation}
where $\mathcal{G}$ is the graph artifact store and $\Sigma$ is the set of community summaries.

Before retrieval, a safety/domain classifier routes the request into one of three classes: crisis, out-of-scope, or in-scope.

\subsection{Retrieval Mechanism}

\textbf{Graph-first retrieval in production:} In the current live API path, evidence is retrieved from the checkpointed graph state and its provenance maps. The router selects among relation-aware branches:
\begin{itemize}
    \item \textbf{Forward lookup:} condition-centric relation expansion,
    \item \textbf{Reverse symptom lookup:} symptom-to-condition ranking,
    \item \textbf{Comparison lookup:} shared and diverging relations between conditions,
    \item \textbf{Entity lookup:} incoming/outgoing relation view for non-condition entities,
    \item \textbf{Global context:} optional community-summary matching.
\end{itemize}

The runtime can optionally synthesize a final answer with an LLM, but only after graph evidence is assembled. Vector retrieval and Neo4j querying exist in the repository for benchmarking and auxiliary workflows; they are not the primary live query backend in the current implementation.

\subsection{Hallucination Reduction Objective}

We define hallucination as any answer claim not supported by retrieved graph facts, relation metadata, or attached source chunk evidence. Let $\mathcal{U}(A)$ be the set of atomic claims in answer $A$, and let $\mathcal{E}$ be the retrieved evidence set. A claim $u \in \mathcal{U}(A)$ is grounded if it is entailed by some $e \in \mathcal{E}$.

The practical objective is:
\begin{equation}
\max \ \text{Accuracy}(A) \quad \text{and} \quad \min \ \mathcal{H}(A)
\end{equation}
with
\begin{equation}
\mathcal{H}(A) = 1 - \frac{|\mathcal{U}_{\text{grounded}}(A)|}{|\mathcal{U}(A)|}
\end{equation}

The implemented system reduces $\mathcal{H}(A)$ through:
\begin{itemize}
    \item schema-constrained extraction and strict triplet validation,
    \item post-extraction cleaning and repair stages,
    \item graph-first retrieval with explicit citation and source-chunk return,
    \item fallback behavior when evidence is insufficient,
    \item crisis short-circuit logic that prevents normal QA generation for high-risk inputs.
\end{itemize}

This formulation matches the current codebase objective: evidence-grounded, safety-aware clinical responses with transparent provenance.


\end{document}